---
title: "Levas Flor Occupancy"
author: "Kaitlyn Gaynor & Lindsey Rich"
date: "3/23/20"
output: html_document
editor_options: 
  chunk_output_type: console
---

Multi-species occupancy modeling

Rather than re-run Tara's code, using my code and adapting for Tara's data.

Get set-up. Load libraries, etc.
```{r setup}
library(R2jags)
library(reshape)
library(reshape2)
library(plyr)
library(dplyr)
library(here)
library(tidyverse)
library(xlsx)
```

Define function for standardizing the covariates. You'll need this later.
```{r stdev}
zstd <- function (x)  (x-mean(x,na.rm = T))/sd(x,na.rm = T)  
```

# Import and manipulate detection/operation histories and metadata.

```{r import data}
D <- read.xlsx("Data2.xlsx", sheetName = "daily_occ")

psi_CoVar <- read.xlsx("Data2.xlsx", sheetName = "PsiCov")
det_CoVar <- read.xlsx("Data2.xlsx", sheetName = "DetCov")
Covariates <- join(psi_CoVar, det_CoVar)
names(Covariates) # take a look at covariate names

# Species IDs and traits (reduce to species present in LF)
Spp <- read.csv("~/Documents/github-repos/gorongosa/gorongosa-anim-cons/Species_metadata_allsites.csv") %>% 
  filter(Levas_Flor == 1) %>% 
  droplevels()
```

Save species names for use later when exporting model results
```{r save species names}
# sppnames2 is just missing 'SppCode' line
sppnames <- c("SppCode", as.character(unique(Spp$SppCode)))
sppnames2 <- as.character(unique(Spp$SppCode))
```

# Run model

This model includes all six covariates from original analysis.
```{r dry season both years with covariates}
# Reorganize detection dataframe, so that it has one column for species and one for detections (rather than a column for each species)
D <- melt(D,vars = c("StudySite"))
colnames(D)[2:3] = c("Easter_name","Detections")
head(D) # make sure it worked

# Combine detection history with StudySite covariates and species traits
D <- merge(D, Covariates, by = "StudySite") # joins by StudySite
D <- merge(D, Spp, by = "Easter_name") # joins by SppCode
head(D)

# order by SppCode and Study Site
D <- D[order(D$SppCode, D$StudySite),]

# for later group-level analysis
G <- cbind(as.numeric(D$FunctionalGroupBG=="Carnivore"),
           as.numeric(D$FunctionalGroupBG=="Other_forager"),
           as.numeric(D$FunctionalGroupBG=="Primate"),
           as.numeric(D$FunctionalGroupBG=="UngulateB")) 


# select occupancy covariates
X = dplyr::select(D, dist_settle, dist_allriv, ndvi_500buff, dist_ALLactrds)
# normalize them
X = apply(X,2,zstd)

## select detection covariates
dX = dplyr::select(D, human_rate, ndvi)
# normalize them
dX = apply(dX,2,zstd)

## identify group covariates
XG = cbind(X*G[,1], X*G[,2], X*G[,3], X*G[,4])

# Define data and parameters. Load all the data including the detection array, number of sampling occasions, individual species sampled, total number of sampled species, and covariate information.
data <- list(D = D$Detections, # number of detections
             N = ceiling(D[,"num.nights"]), # number of trap-nights for each StudySite
             Species = as.numeric(D$SppCode), # species
             n = nrow(D), 
             nspp = max(as.numeric(D$SppCode)),
             X = X, 
             XG = XG, 
             dX = dX)

# specify the initial values
inits = function() {list(Z = as.numeric(data$D>0))}

# specify the parameters to be monitored
params = c("rho","pbeta","spbeta","sigpbeta","mbeta","sigbeta","sbeta","gbeta","psi.mean","sigma.occ","p.mean","sigma.p","alpha","Z","P")

nc = 3       # number of chains
ni = 60000   # number of iterations
nb = 10000   # burn-in period
nthin = 50   # thinning rate


# Run occupancy model. The model file must be in the correct folder (see script AllMammals_pcov.txt)
out3 <- jags(data = data, 
             inits = inits, 
             parameters.to.save = params, 
             model.file ="AllMammals_pcov.txt", 
             n.chains = nc, 
             n.iter = ni,
             n.burnin = nb, 
             n.thin = nthin)

out3.sum <- out3$BUGSoutput$summary

```

# Export the results

### Occupancy and detection probabilities
```{r}
write.table(x=out3.sum,file="results/fxnlgrpbg-070820.csv",sep=",") # save output

#Name this output alpha and P.
alpha <- out3$BUGSoutput$sims.list$alpha
p <- out3$BUGSoutput$sims.list$P

# we are log transforming alphas b/c they are on logit scale.
expit <- function(x)  1/(1+exp(-x))
logit.alpha <- expit(alpha)
logit.p <- expit(p)

# Columns represent species, so we take the mean of each column to get the mean psi and p value (when covariates are at their mean values) for each species.

psimeans <- colMeans(logit.alpha)
names(psimeans) <- sppnames2
psimeans <- as.data.frame(psimeans)
write.table(x = psimeans, file = "results/fxnlgrpbg-070820_alphaspsi()p().csv", sep = ",")

pmeans <- colMeans(logit.p)
names(pmeans) <- sppnames2
write.table(x = pmeans, file = "results/fxnlgrpbg-070820_detection.csv", sep=",")

# Get the quantiles and 95% confidence intervals for psi and p.

apply(logit.alpha, 2, function(x) sort(x)[])
psiCI <- apply(logit.alpha, 2, function(x) quantile(x,probs = c(0.025,0.1,0.5,0.9,0.975)))
colnames(psiCI) <- sppnames2
write.table(x=psiCI, file="results/fxnlgrpbg-070820_alphaCI.psi()p().csv",sep=",")

apply(logit.p, 2, function(x) sort(x)[])
pCI <- apply(logit.p, 2, function(x) quantile(x,probs = c(0.025,0.1,0.5,0.9,0.975)))
colnames(pCI) <- sppnames2
write.table(x = pCI, file="results/fxnlgrpbg-070820_pCI.psi()p().csv", sep = ",")
```


### Occupancy covariate effects
```{r}
# Define the occupancy covariate effects where mbeta is the community-level hyperparameter, gbeta is the group-level hyperparameter, and sbeta is the species-specific parameter.
mbeta <- out3$BUGSoutput$sims.list$mbeta
gbeta <- out3$BUGSoutput$sims.list$gbeta
sbeta <- out3$BUGSoutput$sims.list$sbeta

# Calculate group-level estimates.
covs <- colnames(X) # define covariates
sizes <- c("UngulateG", "Carnivore", "Other_forager", "Primate", "UnglateB") # define groups

group <- data.frame(expand.grid(covs, sizes), matrix(NA, length(covs) * length(sizes), 4)) # create data frame where number of rows is equal to the number of covariates * the number of groups
colnames(group) <- c("Factor", "Group", "Mean", "SD", "LCI", "UCI")

# Create a loop estimating the reference group values.
for (a in 1:length(covs)){
  group[a,3:6] <- c(mean(mbeta[,a]),sd(mbeta[,a]),quantile(mbeta[,a],c(0.025,0.975)))
}

# Create a second loop estimating the other group values.
for (a in 1:length(covs)){
  for (b in 1:(length(sizes)-1)){
    sims <- mbeta[,a] + gbeta[,((b-1)*ncol(X)+a)]
    group[(ncol(X)*(b)+a),3:6] <- c(mean(sims),sd(sims),quantile(sims,c(0.025,0.975)))
  }
}

# Export table with group values.
write.table(x = group, file = "results/fxnlgrpbg-070820_group.csv", sep = ",", row.names=F)


# calculate distributions for reference group

comm_param <- as.data.frame(mbeta)
names(comm_param) <- covs
head(comm_param)
comm_param_long <- comm_param %>% 
  pivot_longer(cols = everything(), names_to = "Factor", values_to = "Value") 

# and for group hyperparameters
head(gbeta)
group_param <- as.data.frame(gbeta)

group_param_ungg <- comm_param # reference group
names(group_param_ungg) <- covs
group_param_ungg_long <- group_param_ungg %>% 
  pivot_longer(cols = everything(), names_to = "Factor", values_to = "Value") %>% 
  mutate(Group = "UngulateG")

group_param_carn <- group_param[,1:4]
names(group_param_carn) <- covs
group_param_carn_long <- (group_param_carn + group_param_ungg) %>% 
  pivot_longer(cols = everything(), names_to = "Factor", values_to = "Value") %>% 
  mutate(Group = "Carnivore")

group_param_other <- group_param[,5:8]
names(group_param_other) <- covs
group_param_other_long <- (group_param_other + group_param_ungg) %>% 
  pivot_longer(cols = everything(), names_to = "Factor", values_to = "Value") %>% 
  mutate(Group = "Other_forager")

group_param_prim <- group_param[,9:12]
names(group_param_prim) <- covs
group_param_prim_long <- (group_param_prim + group_param_ungg) %>% 
  pivot_longer(cols = everything(), names_to = "Factor", values_to = "Value") %>% 
  mutate(Group = "Primate")

group_param_ungb <- group_param[,13:16]
names(group_param_ungb) <- covs
group_param_ungb_long <- (group_param_ungb + group_param_ungg) %>% 
  pivot_longer(cols = everything(), names_to = "Factor", values_to = "Value") %>% 
  mutate(Group = "UngulateB")

group_param_long <- rbind(group_param_carn_long, group_param_other_long, group_param_prim_long, group_param_ungb_long, group_param_ungg_long)
head(group_param_long)
write.csv(group_param_long, "results/fxnlgrpbg-070820_groupcovposteriors.csv", row.names = F)


# Species level estimates 

# Define the species
spec <- Spp$SppCode

# Define the group levels
levels(Spp$FunctionalGroupBGTest) <- levels(Spp$FunctionalGroupBGTest)[c(1,2,3,4,5)]
gg <- as.numeric(Spp$FunctionalGroupBGTest)

# Define the occupancy covariates and groups
covs <- colnames(X)
sizes <- c("UngulateG", "Carnivore", "Other_forager", "Primate", "UnglateB") # define groups

# Create a data frame where the number of rows is equal to the number of covariates * the number of species
species <- data.frame(expand.grid(covs,spec), matrix(NA,length(covs)*length(spec),4))
colnames(species) <- c("Factor","Species","Mean","SD","LCI","UCI")

# Re-define gbeta
gbeta <- cbind(gbeta, matrix(0,nrow(gbeta),length(covs)))

# Create a loop that will estimate species-specific values for each of the covariates
for (a in 1:length(covs)){
  for (b in 1:length(spec)){
    sims <- mbeta[,a] + gbeta[,((gg[b]-1)*ncol(X)+a)] + sbeta[,b,a]
    species[(ncol(X)*(b-1)+a),3:6] <- c(mean(sims),sd(sims),quantile(sims,c(0.025,0.975)))
  }
}

# export the table
write.table(x=species,file="results/fxnlgrpbg-070820_species.csv",sep=",", row.names=F) 

```


## Species and group richness
```{r}
# Species richness for each site

# Define the z matrix
z = out3$BUGSoutput$sims.list$Z

# Sort the data frame based on species, study site, and diet category
d <- sort_df(data = merge(data.frame(ID = 1:nrow(D),
                                     select(D, SppCode, StudySite)), 
                          data.frame(SppCode = spec, 
                                     Group = Spp$FunctionalGroupBG)
                          ),
             "ID")[,c(1,3,4)]

# Create a new data frame
dz <- data.frame(d,t(z))

# Melt the data frame for easy casting
m.dz <- melt(dz,id.vars = c("SppCode","StudySite","Group") )

# Aggregate the data by summing the values in the z matrix for each StudySite station during each iteration
z.all <- acast(m.dz,StudySite ~ variable, fun.aggregate = sum)
write.table(x=z.all,file="results/fxnlgrpbg-070820_rawrichnessdistribution.csv",sep=",", row.names=F)


# Use the aggregated values to create probability distributions and estimate mean, sd, and 95% credible interval values for StudySite-station specific species richness
z.all <- t(apply(z.all,1,function(x) c(mean(x),sd(x),quantile(x,c(0.025,0.975)))))
names <- rownames(z.all)
rownames(z.all) <- NULL
z.all <- cbind(names,z.all)
colnames(z.all) = c("StudySite", "Mean","SD","LCI","UCI")

# Export estimates of species richness as a table
write.table(x=z.all,file="results/fxnlgrpbg-070820_spprich.csv",sep=",", row.names=F)

# To estimate group richness for each site:

# Aggregate the data by summing the group-specific values in teh z matrix for each StudySite station during each iteration
z.group <- acast(m.dz,StudySite + Group ~ variable, fun.aggregate = sum)

# Use the aggregated values to create probability distributions representing estimated StudySite-station specific group richness
z.group <- t(apply(z.group,1,function(x) c(mean(x),sd(x),quantile(x,c(0.025,0.975)))))
names <- rownames(z.group)
rownames(z.group) <- NULL
z.group <- cbind(names,z.group)
colnames(z.group) = c("StudySite", "Mean","SD","LCI","UCI")

# Export estimates of group richness as a table
write.table(x=z.group,file="results/fxnlgrpbg-070820_grouprich.csv",sep=",", row.names=F)
```

### Calculate Hill numbers
(adapted from Broms et al 2018 https://besjournals.onlinelibrary.wiley.com/doi/10.1111/2041-210X.12296)

```{r}
hill1 <- vector("numeric")
hill2 <- vector("numeric")

for(i in 1:nrow(logit.alpha)) {
  
  sum.alpha <- rowSums(logit.alpha)[i] # add up all occupancy probabilities for all species, in that iteration
  
  hill1.input <- vector("numeric")
  hill2.input <- vector("numeric")
  
  for(j in 1:ncol(logit.alpha)) {
    relative.alpha <- logit.alpha[i,j] / sum.alpha
    hill1.input[j] <- relative.alpha * log(relative.alpha)
    hill2.input[j] <- relative.alpha * relative.alpha
  }

  # Hill1 - represents Shannon diversity
  hill1[i] <- exp(-1 * sum(hill1.input))
  
  # Hill2 - represents Simpson diversity
  hill2[i] <- 1/sum(hill2.input)
}

hill <- cbind(hill1, hill2)


# for ungulates

# make list of only ungulates
spp_key <- Spp %>% 
  filter(FunctionalGroupBG %in% c("UngulateB", "UngulateG")) %>% 
  select(SppCode) %>% 
  as.list() %>% 
  unlist(use.names=FALSE)

# subset alphas for ungulates
logit.alpha.df <- as.data.frame(logit.alpha)
names(logit.alpha.df) <- sppnames2
logit.alpha.ungulate <- logit.alpha.df %>% 
  select(spp_key)

hill1_ungulate <- vector("numeric")
hill2_ungulate <- vector("numeric")

for(i in 1:nrow(logit.alpha.ungulate)) {
  
  sum.alpha <- rowSums(logit.alpha.ungulate)[i] # add up all occupancy probabilities for all species, in that iteration
  
  hill1.input.ungulate <- vector("numeric")
  hill2.input.ungulate <- vector("numeric")
  
  for(j in 1:ncol(logit.alpha.ungulate)) {
    relative.alpha <- logit.alpha.ungulate[i,j] / sum.alpha
    hill1.input.ungulate[j] <- relative.alpha * log(relative.alpha)
    hill2.input.ungulate[j] <- relative.alpha * relative.alpha
  }

  # Hill1 - represents Shannon diversity
  hill1_ungulate[i] <- exp(-1 * sum(hill1.input.ungulate))
  
  # Hill2 - represents Simpson diversity
  hill2_ungulate[i] <- 1/sum(hill2.input.ungulate)
}

hill_ungulate <- cbind(hill1_ungulate, hill2_ungulate)

hill_all <- cbind(hill, hill_ungulate)

write.table(hill_all, "results/fxnlgrpbg-070820_hill.csv", sep = ",", row.names=F)

```